---
title: "Where Title VI Fails"
author: "Uma Gaffney, Cecilia Mendez, and Belise Bwiza"
date: "Last updated on `r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: kable
    code_folding: hide
---
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Set R Markdown chunk defaults:
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE, 
  fig.width = 16/2, fig.height = 9/2
)
```

```{r}
#load packages
library(tidyverse)
library(readr)
library(leaflet)
#load data
SCH_truncated <- read_csv("SCH_truncated.csv")
CRDCdistricts <- read_csv("CRDC2013_14_LEA.csv")
rla20112012 <- read_csv("rla20112012.csv")

#data wrangling: combined male and female data for each race
LEA_demo <- SCH_truncated%>% 
  select(1:5, 8:10, 27:ncol(SCH_truncated)) %>%
  mutate(HI = SCH_ENR_HI_M + SCH_ENR_HI_F) %>%
  mutate(AM = SCH_ENR_AM_M + SCH_ENR_AM_F) %>%
  mutate(BL = SCH_ENR_BL_M + SCH_ENR_BL_F) %>%
  mutate(HP = SCH_ENR_HP_M + SCH_ENR_HP_F) %>%
  mutate(TOT = TOT_ENR_M + TOT_ENR_F) %>%
  #summarize by district
  group_by(LEAID) %>%
  summarize(HI = sum(HI), 
            AM = sum(AM), 
            BL = sum(BL), 
            HP = sum(HP), 
            TOT = sum(TOT)) %>%
  #add variable for what percent minority (- asian) and whether it is greater than 75%
  mutate(minprop = (HI + AM + BL + HP) / TOT) %>%
  #removing one weird outlier that claims to be 400% minority
  filter(minprop <= 1) %>%
  mutate(mindis = ifelse(minprop > .75, "Segregated", "Not Segregated"))

#The LEAID variable for CRDCdistricts and rla20112012 start with a leading 0
#follwoing code removes the 0
CRDCdistricts$LEAID <- as.integer(CRDCdistricts$LEAID)
rla20112012$LEAID <- as.integer(rla20112012$LEAID)


#make the scores numeric; removes unneccessary text from some values 
rla20112012$ALL_RLA00pctprof_1112 <- parse_number(rla20112012$ALL_RLA00pctprof_1112)
rla20112012$MWH_RLA00pctprof_1112 <- parse_number(rla20112012$MWH_RLA00pctprof_1112)
rla20112012$MHI_RLA00pctprof_1112 <- parse_number(rla20112012$MHI_RLA00pctprof_1112)
rla20112012$MBL_RLA00pctprof_1112 <- parse_number(rla20112012$MBL_RLA00pctprof_1112)

#join district civil rights data to wrangled district demographic data and district scores data
joined_districts <- CRDCdistricts %>%
  left_join(LEA_demo, by = "LEAID") %>%
  left_join(rla20112012, by = "LEAID")
```

# Segregated Districts in America

According to the Department of Education's Civil Rights data, approximately 20,000 of the nearly 100,000 schools that accept federal funding are more than 75% underprivileged minorities. Of those 20,000 schools, almost 13,000 of them, or about 65%, are in districts that are also more than 75% minority. Because the schools in these districts are generally similar in terms of ethnic breakdown, less than 400 segregated schools and just 11 segregated districts have a formal desegregation order or plan in place. Yet the remaining districts are in need of just as much or more help.

## Student Outcomes and District Demographics

Although we have established that many school districts in this country are heavily segregated, we have not yet established how this is a problem. Instead of looking at whether separate educational systems could ever be philosophically equal, we examined whether they were equal in terms of student outcomes. The plot below suggests that they are not. This graph shows for each school district in the country what percentage of students were considered proficient in reading and language skills based on their state’s assessment, compared to what percentage of the school district is minorities (In this study, we are considering black, Hispanic, Native American, and Pacific Islander students to be minorities, but not Asian students). Also marked are the few districts with a formal desegregation plan or order in place. As can be seen, the clear majority of districts with nearly 100% minority population have no formal desegregation order in place. A clear trend shows that as the district becomes less and less white, reading proficiency decreases. Therefore, we see that segregated districts are not performing equally to white districts and are not an acceptable option.

```{r}

#scatterplot showing overall reading proficiency by what % district is black or hispanic
joined_districts %>%
  filter(LEA_DESEGPLAN != "-5") %>%
  ggplot(aes(x = minprop, y = ALL_RLA00pctprof_1112 / 100, color = LEA_DESEGPLAN)) + 
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  geom_point(size = 1) + 
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  labs(x = "Underprivileged Minority District Enrollment",
       y = "Percentage of Students at Proficiency Level in Reading", 
       color = "Formal Desegregation Plan",
       title = "Reading Proficiency Relative to Minority Enrollment in Districts")


```

## Segregated Districts Across the Country

One limitation of the above graph is that each state has its own testing method and standards, and they are very difficult to compare across states. As a result, we looked specifically at the distributions for reading proficiency in three large states across different regions in this country, and the results were astounding. In New York, and California, reading proficiency in segregated districts (defined as districts with 75% or greater minority enrollment) is so much lower than unsegregated districts that the middle 50% of the distributions barely overlap. High reading proficiencies in Texas suggest that testing may be different in that state, but even so, the gap between unsegregated and segregated districts is obvious.

```{r}
#boxplot data for three states of interest
#plot scores by segregated status and facet by state
joined_districts %>%
  filter(LEA_STATE == "TX" | LEA_STATE == "NY" | LEA_STATE == "CA") %>%
  ggplot(aes(x = mindis, fill = mindis, y = ALL_RLA00pctprof_1112 / 100)) + 
  geom_boxplot() + 
  facet_wrap(~LEA_STATE) + 
  theme_minimal() + 
  scale_fill_brewer(palette = "Set3") +
  labs(x = "District Enrollment Over 75% Underprivileged Minority (Segregated)", 
       y = "Percentage of Students at Proficiency Level in Reading", 
       fill = "", 
       title = "Relationship between District Segregation and Student Reading Level in Three States") +
  scale_y_continuous(labels = scales::percent)
```


## Who Hurts The Most

Education inequality has skyrocketed over the years in this country. The Stanford Center for Education Policy Analysis released a study which found out that between 1990 and 2010, income segregation between school districts has increased by 15%.^[  Owens, Ann, et al. “Income Segregation Between Schools and School Districts.” American Educational Research Journal, vol. 53, no. 4, Aug. 2016, pp. 1159–1197, doi:10.3102/0002831216652722.] It is easy to look at the above graphics and think that only income, not race, affects the education of these students. However, the graphic below suggests otherwise. 
The following boxplot shows the reading proficiency distribution for black, white, and Hispanic students in either segregated or unsegregated districts. As expected based on the prior analyses, all scores go down somewhat in segregated districts. Although the top half of black and Hispanic students do about the same or even a little better in segregated districts, the bottom 50% do notably worse. What is important to see, however, is that that although white students do much worse in segregated districts than in unsegregated, they still do better than their black and Hispanic classmates. These data suggest that even in the same district, white students are still somehow receiving a better education. 



```{r}
#gather data by ethnicity so that scores can be compared across race
#most rows will be empty (since many districts don't have black or hispanic students)
joined_districts %>%
  select(LEAID, 
         mindis, 
         ALL_RLA00pctprof_1112, 
         MWH_RLA00pctprof_1112, 
         MBL_RLA00pctprof_1112, 
         MHI_RLA00pctprof_1112) %>%
  gather(key = "ethnicity", value = "RLA00pctprof_1112", 3:6) %>%
  #plot this data race vs scores faceting by segregation
  filter(mindis != "NA") %>%
  ggplot(aes(x = ethnicity, fill = ethnicity, y = RLA00pctprof_1112 / 100)) + 
  facet_wrap(~mindis) + 
  geom_boxplot() + 
  labs( y = "Percentage of Students at Proficiency Level in Reading", 
        fill = "Student Race / Ethnicity", 
        x = "", 
        title = "Reading Ability of Students by Race in Segregated or Not Segregated Districts") + 
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Set2", 
                    labels = c("All",  "Black", "Hispanic", "White")) +
  theme_minimal() +
  theme(axis.text.x = element_blank())

```

# Redefining Title VI
Half a century ago, Title VI made it illegal to discriminate against students in public schools on the basis of race or ethnicity. How is it possible, then, that the system that was created in a post-Title VI world still allows nearly 8.5 million students to attend school in segregated, underperforming districts? Other influential factors like geography and income have stratified our educational system, and neither of those will be easy to overcome. The first priority must be to aid struggling schools in these districts so that no children are penalized for where they were born or who their parents are. In the long term, Title VI must evolve so that, wherever possible, integration will cross district lines and students will be welcomed into diverse, thriving schools in districts all over the country. 